---
title: "Analysis of Factors Influencing Household Size in the Philippines"
author: "Group 5"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

# Data Tidy 
```{r}
#| label: libraries
library(tidyverse)    
library(MASS)         
library(AER)         
library(caret)        
library(knitr)        
library(performance)  
library(ggplot2)      
library(dplyr)
library(gt)
library(gridExtra)
```

```{r}
#| label: data
data <- read.csv("C:/Users/25094/Desktop/skills/project2/dataset05.csv", stringsAsFactors = FALSE)
```

```{r}
#| label: data tidy
data <- data %>%
  mutate(
    Region = as.factor(Region),
    Household.Head.Sex = as.factor(Household.Head.Sex),
    Type.of.Household = as.factor(Type.of.Household),
    Electricity = as.factor(Electricity)
  )
data <- data %>%
  mutate(
    Total.Household.Income = as.numeric(Total.Household.Income),
    Total.Food.Expenditure = as.numeric(Total.Food.Expenditure),
    Household.Head.Age = as.numeric(Household.Head.Age),
    Total.Number.of.Family.members = as.numeric(Total.Number.of.Family.members),
    House.Floor.Area = as.numeric(House.Floor.Area),
    House.Age = as.numeric(House.Age),
    Number.of.bedrooms = as.numeric(Number.of.bedrooms)
  )
summary(data)

cat("Number of missing values:", sum(is.na(data)), "\n")
```


# Introduction {#sec-Intro}
The Family Income and Expenditure Survey (FIES) is a nationwide survey conducted every three years in the Philippines, providing critical insights into household income, expenditures, and living conditions. Understanding the factors that influence household size is essential for policymakers, urban planners, and social welfare programs, as it impacts housing demand, resource allocation, and economic policies.

In this study, we aim to identify key household-related variables that influence the number of family members living in a household. We will analyze factors such as household income, food expenditure, house characteristics, and household type to determine their impact on family size.To achieve this, we will employ a Generalized Linear Model (GLM) with a Poisson distribution. This approach will allow us to examine the relationship between household characteristics and family size while accounting for the discrete nature of the dependent variable.


# Exploratory Data Analysis {#sec-EDA}
```{r}
#| fig-cap: Distribution of household size
#| label: fig-1
ggplot(data, aes(x = Total.Number.of.Family.members)) +
  geom_histogram(bins = 15, fill = "blue", color = "black") +
  ggtitle("Distribution of Household Size") +
  xlab("Number of Family Members") +
  ylab("Frequency")
```
The histogram in @fig-1, the target variable distribution plot, exhibits a right-skewed distribution. Most households have 3 to 5 members, while the right tail (where the number of family members exceeds 6) is relatively long, which means that large households do exist but the number is significantly smaller.There are also outliers, specifically households with more than 12 members, but these are quite rare.
This is similar to the Poisson distribution and conforms to the count variable characteristics assumed by the generalized linear model (GLM).

```{r}
#| label: tbl-mean-variance
#| tbl-cap: Summary statistics of household size (mean, variance, and variance-to-mean ratio)

mean_val <- mean(data$Total.Number.of.Family.members, na.rm = TRUE)
var_val <- var(data$Total.Number.of.Family.members, na.rm = TRUE)
variance_mean_ratio <- var_val / mean_val

summary_table <- data.frame(
  Statistic = c("Mean", "Variance", "Variance/Mean Ratio"),
  Value = c(mean_val, var_val, variance_mean_ratio)
)

summary_table |> 
  gt() |>
  fmt_number(columns = c("Value"), decimals = 2) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```
From @tbl-mean-variance, the mean (4.55) and variance (4.79) are very close, and the Poisson distribution is a suitable modeling choice.
Also, the variance/mean ratio (Variance/Mean Ratio) = 1.05, which is close to 1, suggesting that the data does not have obvious overdispersion problems. Therefore, using a Poisson regression model is appropriate.  



The three box plots show the relationship between the number of family members and different categorical variables (household head's gender, household type, electricity supply).
```{r} 
#| fig-cap: Relationship between household head's sex and household size
#| label: fig-2
plot_household_sex<-ggplot(data, aes(x = Household.Head.Sex, y = Total.Number.of.Family.members, fill = Household.Head.Sex)) + 
  geom_boxplot() +
  ggtitle("Household Size by Household Head's Sex") +
  xlab("Household Head's Sex") +
  ylab("Number of Family Members")+
  theme(legend.position = "none")
print(plot_household_sex)
```
@fig-2 presents the distribution of the number of family members of male and female household heads. It shows that male household heads are more likely to have larger families, while female household heads tend to have a more stable household size with fewer extreme large-family cases.

```{r} 
#| fig-cap: Relationship between household type and household size
#| label: fig-3
plot_household_type<-ggplot(data, aes(x = Type.of.Household, y = Total.Number.of.Family.members, fill = Type.of.Household)) + 
  geom_boxplot() +
  ggtitle("Household Size by Type of Household") +
  xlab("Type of Household") +
  ylab("Number of Family Members") +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))+
  theme(legend.position = "none")
print(plot_household_type)
```
@fig-3 illustrates the difference in the number of different types of families. Single families have the smallest household size, mostly ranging between 3 to 5 members, but there are outliers with more than eight members. Extended families and Two or More Nonrelated Persons have a larger number of family members, with higher medians and interquartile ranges (IQR). However, there are more outliers in extended families, and some have more than 12 family members.
Extended families have significantly more members because they may include grandparents, siblings, and other relatives. Two or More Nonrelated Persons refers to individuals who live together but have no direct family relationship. These people may share the same residence, but are not linked by blood, marriage, or adoption. It is an arrangement based on shared housing and economic resources, rather than kinship.

```{r} 
#| fig-cap: Relationship between electricity supply and household size
#| label: fig-4
plot_electricity<-ggplot(data, aes(x = Electricity, y = Total.Number.of.Family.members)) +
  geom_boxplot() +
  labs(title = "Household Size by Electricity Access",
       x = "Has Electricity (1=Yes, 0=No)",
       y = "Number of Family Members") 
print(plot_electricity)
```
@fig-4 shows that the direct impact of electricity supply on household size may be small, as the medians of the two sets of data are almost the same, with a household size of about 4 members. However, households with electricity display a slightly wider interquartile range (IQR), indicating more variation in household sizes.Outliers (households with more than 10 members) appear in both households without electricity and households with electricity, and it is possible that some large families still live without electricity.


The following section will analyze the relationship between continuous variables and the number of people living in a household.
```{r}
#| fig-cap: Relationship between household income and household size
#| label: fig-5
income_boxplot <- data %>%
  mutate(Income.Group = cut(Total.Household.Income, 
                           breaks = quantile(Total.Household.Income, probs = seq(0, 1, by = 0.25)), labels = c("Low Income", "Lower Middle", "Upper Middle", "High Income"), include.lowest = TRUE)) %>%
  ggplot(aes(x = Income.Group, y = Total.Number.of.Family.members, fill = Income.Group)) +
  geom_boxplot() +
  labs(title = "Household Size by Income Level",
       x = "Income Group",
       y = "Number of Family Members") +
  theme(legend.position = "none")
print(income_boxplot)
```
In @fig-5, households with low income tend to have smaller family sizes, with a median below three members, although some households still have more than 8 members.For the lower-middle income group, the median household size is around 5, with a few extreme values.In both the upper-middle and high-income groups, the median also sits around 5 members, but these groups exhibit more extreme cases, including households with over 12 members.

```{r}
#| fig-cap: Relationship between food expenditure and household size
#| label: fig-6
food_exp_by_family_size <- data %>%
  mutate(Food.Exp.Group = cut(Total.Food.Expenditure, 
                              breaks = quantile(Total.Food.Expenditure, probs = seq(0, 1, by = 0.2), na.rm = TRUE), 
                              labels = c("Very Low", "Low", "Medium", "High", "Very High"),
                              include.lowest = TRUE)) %>%
  ggplot(aes(x = Food.Exp.Group, y = Total.Number.of.Family.members, fill = Food.Exp.Group)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Household Size by Food Expenditure Level",
       x = "Food Expenditure Level",
       y = "Household Size (Number of Family Members)") +
  scale_y_continuous(breaks = seq(0, max(data$Total.Number.of.Family.members, na.rm = TRUE), by = 2)) +
  theme(legend.position = "none")

print(food_exp_by_family_size)
```
In @fig-6, as food expenditure increases, household size shows an upward trend, and the number of outliers also increases.Food expenditure is grouped by 20% quantile into "Very Low", "Low", "Medium", "High", "Very High". Households in the "Very Low" expenditure group have a more concentrated household size, mostly ranging from 2 to 4 members. For the "Very High" expenditure group, the median household size rises to around 5–6 members, and the upper quartile is higher, indicating that larger families are more likely to fall into this category. Additionally, there are more higher outliers, suggesting that household sizes vary more greatly among those with higher food expenditures.

```{r}
#| fig-cap: Relationship between household head's age and household size
#| label: fig-7
head_age_line <- data %>% 
  mutate(Age.Group = cut(Household.Head.Age, 
                         breaks = c(0,20, 30, 40, 50, 60, Inf),
                         labels = c("< 20","20-29", "30-39", "40-49", "50-59", ">=60"),
                         include.lowest = TRUE)) %>%
 
  group_by(Age.Group) %>%
  summarise(Mean.Family.Size = mean(Total.Number.of.Family.members, na.rm = TRUE)) %>%

  ggplot(aes(x = Age.Group, y = Mean.Family.Size, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Household Head Age vs Average Household Size",
       x = "Age Group", 
       y = "Average Family Size") 

print(head_age_line)
```
@fig-7 shows an inverted U-shaped trend in household size based on the head's age. Households with heads under 20 are smallest (about 2.5 persons). The size of households headed by heads of households aged 20-49 years old increases with age. The average household size of heads of households aged 20-29 years old is relatively small, about 4 people, and reaches a peak in the 40–49 age group, where the average number of family members is greater than 5.
Younger households tend to be smaller, likely due to fewer children and economic pressure, which may lead young people to establish smaller household units.In contrast, households headed by individuals aged 30–49 reach their peak in size — this group is often more financially stable, may have multiple children, or live with elderly parents.After age 50, household size shrinks as children grow up and move out, and some households may have fewer members due to divorce or widowhood.

```{r}
#| fig-cap: Relationship between the number of bedrooms and household size
#| label: fig-8
ggplot(data, aes(x = as.factor(Number.of.bedrooms), y = Total.Number.of.Family.members, fill = as.factor(Number.of.bedrooms))) + 
  geom_boxplot() +
  ggtitle("Household Size vs. Number of Bedrooms") +
  xlab("Number of Bedrooms") +
  ylab("Number of Family Members") +
  theme(legend.position = "none")
```
In @fig-8, overall, household size increases with the number of bedrooms.
As the number of bedrooms increases from 0 to 6,  the median household size remains relatively stable, between 4 and 6 members, indicating that household size is not strictly dependent on the number of bedrooms. Households with fewer bedrooms may accommodate more members, as seen from the presence of outliers in the 0-4 bedroom groups. In some cases, up to 12 people live together, suggesting room sharing occurs. On the other hand, households with more bedrooms may have fewer members, such as wealthier families or elderly couples.

```{r}
#| fig-cap: Relationship between floor area and household size
#| label: fig-9
floor_area_analysis <- data %>%
  mutate(Floor.Area.Group = cut(House.Floor.Area, 
                                breaks = c(0, 30, 60, 90, 120, 150, 200, Inf),
                                labels = c("0-30", "30-60", "60-90", "90-120", "120-150", "150-200", "200+"),
                                include.lowest = TRUE)) %>%
  group_by(Floor.Area.Group) %>%
  summarise(
    Mean.Family.Size = mean(Total.Number.of.Family.members, na.rm = TRUE),
    Median.Family.Size = median(Total.Number.of.Family.members, na.rm = TRUE),
    Count = n()
  )

floor_area_plot <- ggplot(floor_area_analysis, aes(x = Floor.Area.Group, group = 1)) +
  geom_line(aes(y = Mean.Family.Size), color = "blue", size = 1.2, linetype = "solid") + 
  geom_line(aes(y = Median.Family.Size), color = "red", size = 1.2, linetype = "dashed") + 
  geom_point(aes(y = Mean.Family.Size), color = "blue", size = 3) +
  geom_point(aes(y = Median.Family.Size), color = "red", size = 3) +
  labs(title = "Floor Area vs Family Size",
       x = "Floor Area (sq.m)", 
       y = "Family Size",
       caption = "Blue：Mean，Red：Median") +
  theme_minimal()

floor_area_bar <- ggplot(floor_area_analysis, aes(x = Floor.Area.Group, y = Count)) +
  geom_bar(stat = "identity", fill = "gray", alpha = 0.7) +
  labs(title = "Data Count by Floor Area Group",
       x = "Floor Area (sq.m)",
       y = "Count of Households") +
  theme_minimal()

grid.arrange(floor_area_plot, floor_area_bar, ncol = 1)
```
In @fig-9, the blue solid line represents the average household size, which can be influenced by extreme values. The red dotted line shows the median household size, which is more representative of the typical family. Household size increases as floor area increases, reaching its peak in the 90–120 square meter range. When the floor area exceeds 150 square meters, household size shows a slight decreasing trend. For very small homes (0–30 sqm), the average household size is the smallest, with fewer than 4.5 members. 
However, most households live in smaller homes, especially in the 0–60 sqm group, indicating that small housing units are the most common, and the number of households decreases as the floor area increases.

```{r}
#| fig-cap: Relationship between house age and household size
#| label: fig-10
house_age_analysis <- data %>%
  mutate(House.Age.Group = cut(House.Age, 
                               breaks = c(0,5, 10,15, 20,25, 30,35, 40, 100),
                               labels = c("0-5","5-10", "10-15","15-20", "20-25", "25-30","30-35","35-40","40+"),
                               include.lowest = TRUE)) %>%
  group_by(House.Age.Group) %>%
  summarise(
    Mean.Family.Size = mean(Total.Number.of.Family.members, na.rm = TRUE),
    Median.Family.Size = median(Total.Number.of.Family.members, na.rm = TRUE),
    Count = n()
  )

plot_mean <- ggplot(house_age_analysis, aes(x = House.Age.Group, y = Mean.Family.Size, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  geom_line(aes(y = Median.Family.Size, group = 1), color = "red", linetype = "dashed", size = 1) + 
  geom_point(aes(y = Median.Family.Size), color = "red", size = 3) +
  labs(title = "House Age vs Family Size",
       x = "House Age (years)", 
       y = "Average / Median Family Size") +
  annotate("text", x = 4, y = max(house_age_analysis$Mean.Family.Size) + 0.2, label = "Blue = Mean, Red = Median", color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

plot_count <- ggplot(house_age_analysis, aes(x = House.Age.Group, y = Count)) +
  geom_bar(stat = "identity", fill = "gray", alpha = 0.5) +
  labs(title = "Data Count by House Age Group",
       x = "House Age (years)", 
       y = "Count of Households") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(plot_mean, plot_count, ncol = 1)

```
In @fig-10, due to the influence of extreme values, the mean household size is generally higher than the median.Homes with an age of 0–5 years have larger family sizes (over 4.5 members), suggesting that newer houses may attract larger households. However, there is a secondary peak around 35–40 years, indicating that some large families still choose to live in older homes.For houses aged 10–30 years, the household size remains stable, with an average of around 4.5 members, suggesting that during this period, the impact of house age on family size is relatively small.
Most people tend to live in houses under 20 years old, while homes older than 20 years are associated with smaller household sizes.It may be because some old houses may have been rented out or demolished, resulting in a decrease in family size.





#Formal analytics


