
---
title: "Analysis of Factors Influencing Household Size in the Philippines"
author: "Group 5"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  eval: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

# Data Tidy

```{r}
#| label: libraries
library(tidyverse)    
library(MASS)         
library(AER)         
library(caret)        
library(knitr)        
library(performance)  
library(ggplot2)      
library(dplyr)
library(gt)
library(gridExtra)
library(car)
library(pscl)
```

```{r}
#| label: data
data <- read.csv("C:/Users/25094/Desktop/skills/project2/dataset05.csv", stringsAsFactors = FALSE)
```

```{r}
#| label: data tidy
data <- data %>%
  mutate(
    Region = as.factor(Region),
    Household.Head.Sex = as.factor(Household.Head.Sex),
    Type.of.Household = as.factor(Type.of.Household),
    Electricity = as.factor(Electricity)
  )
data <- data %>%
  mutate(
    Total.Household.Income = as.numeric(Total.Household.Income),
    Total.Food.Expenditure = as.numeric(Total.Food.Expenditure),
    Household.Head.Age = as.numeric(Household.Head.Age),
    Total.Number.of.Family.members = as.numeric(Total.Number.of.Family.members),
    House.Floor.Area = as.numeric(House.Floor.Area),
    House.Age = as.numeric(House.Age),
    Number.of.bedrooms = as.numeric(Number.of.bedrooms)
  )
summary(data)

cat("Number of missing values:", sum(is.na(data)), "\n")
```

# Introduction {#sec-Intro}

The Family Income and Expenditure Survey (FIES) is a nationwide survey conducted every three years in the Philippines, providing critical insights into household income, expenditures, and living conditions.
Understanding the factors that influence household size is essential for policymakers, urban planners, and social welfare programs, as it impacts housing demand, resource allocation, and economic policies.

In this study, we aim to identify key household-related variables that influence the number of family members living in a household.
We will analyze factors such as household income, food expenditure, house characteristics, and household type to determine their impact on family size.To achieve this, we will employ a Generalized Linear Model (GLM) with a Poisson distribution.
This approach will allow us to examine the relationship between household characteristics and family size while accounting for the discrete nature of the dependent variable.

# Exploratory Data Analysis {#sec-EDA}

```{r}
#| fig-cap: Distribution of household size
#| label: fig-1
ggplot(data, aes(x = Total.Number.of.Family.members)) +
  geom_histogram(bins = 15, fill = "blue", color = "black") +
  ggtitle("Distribution of Household Size") +
  xlab("Number of Family Members") +
  ylab("Frequency")
```

The histogram in @fig-1, the target variable distribution plot, exhibits a right-skewed distribution.
Most households have 3 to 5 members, while the right tail (where the number of family members exceeds 6) is relatively long, which means that large households do exist but the number is significantly smaller.There are also outliers, specifically households with more than 12 members, but these are quite rare.
This is similar to the Poisson distribution and conforms to the count variable characteristics assumed by the generalized linear model (GLM).

```{r}
#| label: tbl-mean-variance
#| tbl-cap: Summary statistics of household size (mean, variance, and variance-to-mean ratio)

mean_val <- mean(data$Total.Number.of.Family.members, na.rm = TRUE)
var_val <- var(data$Total.Number.of.Family.members, na.rm = TRUE)
variance_mean_ratio <- var_val / mean_val

summary_table <- data.frame(
  Statistic = c("Mean", "Variance", "Variance/Mean Ratio"),
  Value = c(mean_val, var_val, variance_mean_ratio)
)

summary_table |> 
  gt() |>
  fmt_number(columns = c("Value"), decimals = 2) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

From @tbl-mean-variance, the mean (4.55) and variance (4.79) are very close, and the Poisson distribution is a suitable modeling choice.
Also, the variance/mean ratio (Variance/Mean Ratio) = 1.05, which is close to 1, suggesting that the data does not have obvious overdispersion problems.
Therefore, using a Poisson regression model is appropriate.

The three box plots show the relationship between the number of family members and different categorical variables (household head's gender, household type, electricity supply).

```{r}
#| fig-cap: Relationship between household head's sex and household size
#| label: fig-2
plot_household_sex<-ggplot(data, aes(x = Household.Head.Sex, y = Total.Number.of.Family.members, fill = Household.Head.Sex)) + 
  geom_boxplot() +
  ggtitle("Household Size by Household Head's Sex") +
  xlab("Household Head's Sex") +
  ylab("Number of Family Members")+
  theme(legend.position = "none")
print(plot_household_sex)
```

@fig-2 presents the distribution of the number of family members of male and female household heads.
It shows that male household heads are more likely to have larger families, while female household heads tend to have a more stable household size with fewer extreme large-family cases.

```{r}
#| fig-cap: Relationship between household type and household size
#| label: fig-3
plot_household_type<-ggplot(data, aes(x = Type.of.Household, y = Total.Number.of.Family.members, fill = Type.of.Household)) + 
  geom_boxplot() +
  ggtitle("Household Size by Type of Household") +
  xlab("Type of Household") +
  ylab("Number of Family Members") +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))+
  theme(legend.position = "none")
print(plot_household_type)
```

@fig-3 illustrates the difference in the number of different types of families.
Single families have the smallest household size, mostly ranging between 3 to 5 members, but there are outliers with more than eight members.
Extended families and Two or More Nonrelated Persons have a larger number of family members, with higher medians and interquartile ranges (IQR).
However, there are more outliers in extended families, and some have more than 12 family members.
Extended families have significantly more members because they may include grandparents, siblings, and other relatives.
Two or More Nonrelated Persons refers to individuals who live together but have no direct family relationship.
These people may share the same residence, but are not linked by blood, marriage, or adoption.
It is an arrangement based on shared housing and economic resources, rather than kinship.

```{r}
#| fig-cap: Relationship between electricity supply and household size
#| label: fig-4
plot_electricity<-ggplot(data, aes(x = Electricity, y = Total.Number.of.Family.members)) +
  geom_boxplot() +
  labs(title = "Household Size by Electricity Access",
       x = "Has Electricity (1=Yes, 0=No)",
       y = "Number of Family Members") 
print(plot_electricity)
```

@fig-4 shows that the direct impact of electricity supply on household size may be small, as the medians of the two sets of data are almost the same, with a household size of about 4 members.
However, households with electricity display a slightly wider interquartile range (IQR), indicating more variation in household sizes.Outliers (households with more than 10 members) appear in both households without electricity and households with electricity, and it is possible that some large families still live without electricity.

The following section will analyze the relationship between continuous variables and the number of people living in a household.

```{r}
#| fig-cap: Relationship between household income and household size
#| label: fig-5
income_boxplot <- data %>%
  mutate(Income.Group = cut(Total.Household.Income, 
                           breaks = quantile(Total.Household.Income, probs = seq(0, 1, by = 0.25)), labels = c("Low Income", "Lower Middle", "Upper Middle", "High Income"), include.lowest = TRUE)) %>%
  ggplot(aes(x = Income.Group, y = Total.Number.of.Family.members, fill = Income.Group)) +
  geom_boxplot() +
  labs(title = "Household Size by Income Level",
       x = "Income Group",
       y = "Number of Family Members") +
  theme(legend.position = "none")
print(income_boxplot)
```

In @fig-5, households with low income tend to have smaller family sizes, with a median below three members, although some households still have more than 8 members.For the lower-middle income group, the median household size is around 5, with a few extreme values.In both the upper-middle and high-income groups, the median also sits around 5 members, but these groups exhibit more extreme cases, including households with over 12 members.

```{r}
#| fig-cap: Relationship between food expenditure and household size
#| label: fig-6
food_exp_by_family_size <- data %>%
  mutate(Food.Exp.Group = cut(Total.Food.Expenditure, 
                              breaks = quantile(Total.Food.Expenditure, probs = seq(0, 1, by = 0.2), na.rm = TRUE), 
                              labels = c("Very Low", "Low", "Medium", "High", "Very High"),
                              include.lowest = TRUE)) %>%
  ggplot(aes(x = Food.Exp.Group, y = Total.Number.of.Family.members, fill = Food.Exp.Group)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Household Size by Food Expenditure Level",
       x = "Food Expenditure Level",
       y = "Household Size (Number of Family Members)") +
  scale_y_continuous(breaks = seq(0, max(data$Total.Number.of.Family.members, na.rm = TRUE), by = 2)) +
  theme(legend.position = "none")

print(food_exp_by_family_size)
```

In @fig-6, as food expenditure increases, household size shows an upward trend, and the number of outliers also increases.Food expenditure is grouped by 20% quantile into "Very Low", "Low", "Medium", "High", "Very High".
Households in the "Very Low" expenditure group have a more concentrated household size, mostly ranging from 2 to 4 members.
For the "Very High" expenditure group, the median household size rises to around 5–6 members, and the upper quartile is higher, indicating that larger families are more likely to fall into this category.
Additionally, there are more higher outliers, suggesting that household sizes vary more greatly among those with higher food expenditures.

```{r}
#| fig-cap: Relationship between household head's age and household size
#| label: fig-7
head_age_line <- data %>% 
  mutate(Age.Group = cut(Household.Head.Age, 
                         breaks = c(0,20, 30, 40, 50, 60, Inf),
                         labels = c("< 20","20-29", "30-39", "40-49", "50-59", ">=60"),
                         include.lowest = TRUE)) %>%
 
  group_by(Age.Group) %>%
  summarise(Mean.Family.Size = mean(Total.Number.of.Family.members, na.rm = TRUE)) %>%

  ggplot(aes(x = Age.Group, y = Mean.Family.Size, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Household Head Age vs Average Household Size",
       x = "Age Group", 
       y = "Average Family Size") 

print(head_age_line)
```

@fig-7 shows an inverted U-shaped trend in household size based on the head's age.
Households with heads under 20 are smallest (about 2.5 persons).
The size of households headed by heads of households aged 20-49 years old increases with age.
The average household size of heads of households aged 20-29 years old is relatively small, about 4 people, and reaches a peak in the 40–49 age group, where the average number of family members is greater than 5.
Younger households tend to be smaller, likely due to fewer children and economic pressure, which may lead young people to establish smaller household units.In contrast, households headed by individuals aged 30–49 reach their peak in size — this group is often more financially stable, may have multiple children, or live with elderly parents.After age 50, household size shrinks as children grow up and move out, and some households may have fewer members due to divorce or widowhood.

```{r}
#| fig-cap: Relationship between the number of bedrooms and household size
#| label: fig-8
ggplot(data, aes(x = as.factor(Number.of.bedrooms), y = Total.Number.of.Family.members, fill = as.factor(Number.of.bedrooms))) + 
  geom_boxplot() +
  ggtitle("Household Size vs. Number of Bedrooms") +
  xlab("Number of Bedrooms") +
  ylab("Number of Family Members") +
  theme(legend.position = "none")
```

In @fig-8, overall, household size increases with the number of bedrooms.
As the number of bedrooms increases from 0 to 6, the median household size remains relatively stable, between 4 and 6 members, indicating that household size is not strictly dependent on the number of bedrooms.
Households with fewer bedrooms may accommodate more members, as seen from the presence of outliers in the 0-4 bedroom groups.
In some cases, up to 12 people live together, suggesting room sharing occurs.
On the other hand, households with more bedrooms may have fewer members, such as wealthier families or elderly couples.

```{r}
#| fig-cap: Relationship between floor area and household size
#| label: fig-9
floor_area_analysis <- data %>%
  mutate(Floor.Area.Group = cut(House.Floor.Area, 
                                breaks = c(0, 30, 60, 90, 120, 150, 200, Inf),
                                labels = c("0-30", "30-60", "60-90", "90-120", "120-150", "150-200", "200+"),
                                include.lowest = TRUE)) %>%
  group_by(Floor.Area.Group) %>%
  summarise(
    Mean.Family.Size = mean(Total.Number.of.Family.members, na.rm = TRUE),
    Median.Family.Size = median(Total.Number.of.Family.members, na.rm = TRUE),
    Count = n()
  )

floor_area_plot <- ggplot(floor_area_analysis, aes(x = Floor.Area.Group, group = 1)) +
  geom_line(aes(y = Mean.Family.Size), color = "blue", size = 1.2, linetype = "solid") + 
  geom_line(aes(y = Median.Family.Size), color = "red", size = 1.2, linetype = "dashed") + 
  geom_point(aes(y = Mean.Family.Size), color = "blue", size = 3) +
  geom_point(aes(y = Median.Family.Size), color = "red", size = 3) +
  labs(title = "Floor Area vs Family Size",
       x = "Floor Area (sq.m)", 
       y = "Family Size",
       caption = "Blue：Mean，Red：Median") +
  theme_minimal()

floor_area_bar <- ggplot(floor_area_analysis, aes(x = Floor.Area.Group, y = Count)) +
  geom_bar(stat = "identity", fill = "gray", alpha = 0.7) +
  labs(title = "Data Count by Floor Area Group",
       x = "Floor Area (sq.m)",
       y = "Count of Households") +
  theme_minimal()

grid.arrange(floor_area_plot, floor_area_bar, ncol = 1)
```

In @fig-9, the blue solid line represents the average household size, which can be influenced by extreme values.
The red dotted line shows the median household size, which is more representative of the typical family.
Household size increases as floor area increases, reaching its peak in the 90–120 square meter range.
When the floor area exceeds 150 square meters, household size shows a slight decreasing trend.
For very small homes (0–30 sqm), the average household size is the smallest, with fewer than 4.5 members.
However, most households live in smaller homes, especially in the 0–60 sqm group, indicating that small housing units are the most common, and the number of households decreases as the floor area increases.

```{r}
#| fig-cap: Relationship between house age and household size
#| label: fig-10
house_age_analysis <- data %>%
  mutate(House.Age.Group = cut(House.Age, 
                               breaks = c(0,5, 10,15, 20,25, 30,35, 40, 100),
                               labels = c("0-5","5-10", "10-15","15-20", "20-25", "25-30","30-35","35-40","40+"),
                               include.lowest = TRUE)) %>%
  group_by(House.Age.Group) %>%
  summarise(
    Mean.Family.Size = mean(Total.Number.of.Family.members, na.rm = TRUE),
    Median.Family.Size = median(Total.Number.of.Family.members, na.rm = TRUE),
    Count = n()
  )

plot_mean <- ggplot(house_age_analysis, aes(x = House.Age.Group, y = Mean.Family.Size, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  geom_line(aes(y = Median.Family.Size, group = 1), color = "red", linetype = "dashed", size = 1) + 
  geom_point(aes(y = Median.Family.Size), color = "red", size = 3) +
  labs(title = "House Age vs Family Size",
       x = "House Age (years)", 
       y = "Average / Median Family Size") +
  annotate("text", x = 4, y = max(house_age_analysis$Mean.Family.Size) + 0.2, label = "Blue = Mean, Red = Median", color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

plot_count <- ggplot(house_age_analysis, aes(x = House.Age.Group, y = Count)) +
  geom_bar(stat = "identity", fill = "gray", alpha = 0.5) +
  labs(title = "Data Count by House Age Group",
       x = "House Age (years)", 
       y = "Count of Households") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(plot_mean, plot_count, ncol = 1)

```

In @fig-10, due to the influence of extreme values, the mean household size is generally higher than the median.Homes with an age of 0–5 years have larger family sizes (over 4.5 members), suggesting that newer houses may attract larger households.
However, there is a secondary peak around 35–40 years, indicating that some large families still choose to live in older homes.For houses aged 10–30 years, the household size remains stable, with an average of around 4.5 members, suggesting that during this period, the impact of house age on family size is relatively small.
Most people tend to live in houses under 20 years old, while homes older than 20 years are associated with smaller household sizes.It may be because some old houses may have been rented out or demolished, resulting in a decrease in family size.

# Formal analytics {#sec-Formal}

```{r}
#| label: full-poisson
# Poisson Full Model
poisson_full_model <- glm(
  Total.Number.of.Family.members ~
    Total.Household.Income +
    Total.Food.Expenditure +
    Household.Head.Age +
    Number.of.bedrooms +
    House.Floor.Area +
    House.Age +
    Type.of.Household +
    Electricity +
    Household.Head.Sex,
  family = poisson, data = data
)

summary(poisson_full_model)
```

In this study, we modeled the total number of family members based on Poisson's generalized linear model (GLM) to explore the effects of economic, housing, and social structure factors on family size.
Analyzing data from 1,788 households, we found that income, food expenditures, age of the head of household, household type, and gender were the main factors affecting household size, while housing characteristics and access to electricity played a lesser role.

The modeling equations and analysis of which are as follows

$$
\begin{array}{rcl}
\log(E[Y]) & = & \alpha \\
& + & \hat{\beta}_{\text{Income}} \cdot \text{Income} \\
& + & \hat{\beta}_{\text{Food}} \cdot \text{Food Expenditure} \\
& + & \hat{\beta}_{\text{Age}} \cdot \text{Household Head Age} \\
& + & \hat{\beta}_{\text{HouseAge}} \cdot \text{House Age} \\
& + & \hat{\beta}_{\text{Type}} \cdot I(\text{Household Type}) \\
& + & \hat{\beta}_{\text{Sex}} \cdot I(\text{Household Head Sex})
\end{array}
$$

**Where:**

-   $Y$: Total number of family members (dependent variable)\
-   $\log$: Natural logarithm link function\
-   $\hat{\beta}_{*}$: Coefficients estimated from the model\
-   $I(\cdot)$: Indicator function for categorical variables

## Main influencing factors

Regression analysis shows that household income (Total.Household.Income) is negatively correlated with the number of household members, with a regression coefficient of -5.593e-07 (p \< 0.001), indicating that the higher the income, the lower the number of household members\*.
Specifically, for every 1 currency unit increase in household income, the logarithmic expected value of the number of household members decreases by 5.593e-07.
For example, an increase in income from 154,715 to 535,899 is expected to result in a decrease in the number of household members by 0.21.
This suggests that better-off families prefer the nuclear family model (parents living with their children) to multigenerational cohabitation.
Recommendation: Housing or social welfare policies could be geared towards the lower and middle income groups to encourage extended family patterns and improve family stability.

Regression analysis of the effect of food expenditure on the number of household members (Total.Food.Expenditure) shows that the higher the food expenditure, the higher the number of household members (regression coefficient 5.232e-06, p \< 0.001).
The logarithmic expected value of the number of household members increases by 5.232e-06 for every 1 currency unit increase in food expenditure.
e.g., an increase in expenditure from 39,580 to 133,445 is expected to increase the number of household members by 0.49.
This indicates that larger households spend more on food.
This suggests that larger households have higher food expenditures and that food expenditures can be used as a measure of household size.
Recommendation: Provide food subsidies to reduce the cost of living for families with many children.

Age of the head of household on household size (Household.Head.Age) The regression results show that the age of the head of household is negatively correlated with the number of household members (regression coefficient -4.400e-03, p \< 0.001), i.e., the older the head of the household, the smaller the size of the household.
The logarithmic expected value of the number of household members decreases by 4.400e-03 with each additional year of age; for example, a 46-year-old head of household is expected to have 0.11 fewer household members than a 21-year-old head of household.
This trend may be due to the fact that children of older heads of household are living independently, while younger heads of household prefer to live with their families, especially in an environment of greater financial stress.
Recommendation: Improve the social support system for the elderly and provide better community-based elderly and psychosocial care services.

Impact of household type on the number of household members (Type.of.Household) The regression results show that the number of single household members is significantly less than that of extended households (regression coefficient -0.337, p \< 0.001 ).
This reflects the increasing prevalence of solitary or small family structures in modern society.
Recommendation: More small family housing and rental support suitable for people living alone should be provided.

The analysis of the effect of gender of household head on the number of household members (Household.Head.Sex) showed that the number of household members of male household head was significantly more than that of female household head at 52.5% (regression coefficient 0.243, p \< 0.001).
This may be related to the traditional culture where men live with their families more often.
Recommendation: Strengthen welfare support for female heads of household to enhance their independent living security.

## Factors with small or insignificant effects

The results of the (non-significant) effect of housing and electricity on the number of household members show that the size of the house and the number of bedrooms do not have a significant effect on the number of household members (coefficients of -2.660e-04 and -8.741e-03, respectively, are non-significant), suggesting that the size of the household is driven more by economic and cultural factors than by the size of the living space.
Meanwhile, the electricity supply variable is also not significant (coefficient -7.259e-03, p = 0.81), probably due to the lack of a basis for comparison as most of the households in the sample are already electrified.
Recommendations: Housing policies should focus on quality of living rather than size expansion, and infrastructure studies could introduce more differentiated indicators (e.g., water supply, internet).

By observing the model as a whole, it can be seen that the model residual deviation (1242.8) is significantly lower than the original deviation (1854.6), indicating a good fit; the AIC is 7120.1, which has good explanatory power.
Most of the variables are significant and the model is valid, but it can still be further optimized by eliminating insignificant variables or considering negative binomial regression.

```{r}
#| label: stepwise-selection
# Step-by-step variable filtering using step
selected_model1 <- step(poisson_full_model, direction = "backward", trace = TRUE)

summary(selected_model1)

# View the AIC of the filtered model
cat("AIC of the selected model:", AIC(selected_model1), "\n")
```

In the stepwise regression analysis, we screened the variables of the initially constructed Poisson regression model (Poisson Full Model) by using the stepwise elimination method (DIRECTION = "backward"), and used the AIC (Akaikei Information Criterion) as the model preference criterion to search for a model that is more concise and with a higher degree of fit.
The AIC of the original model is 7120.13, and the AIC of the finally selected optimal model is reduced to 7115.55, which indicates that the model successfully eliminates some redundant variables and improves the explanatory efficiency without losing the predictive ability.

The following variables were retained in the final model:

Household.

Food.Expenditure (food expenditure)

Household.Head.Age

House.Age

Type.of.Household

Household.Head.Sex (sex of head of household)

Several variables in the model are highly significant (p \< 0.001), such as Total.Household.Income (coefficient = -5.949e-07, p \< 0.001), Total.Food.Expenditure (coefficient = 5.186e-06, p \< 0.001), Household.
Head.Age (coefficient = -4.512e-03, p \< 0.001), Type.of.HouseholdSingle Family (coefficient = -0.334, p \< 0.001), and Household.Head.SexMale (coefficient = 0.243, p \< 0.001).
These results indicate that household structure, economic expenditures, and demographic characteristics have a significant effect on the number of household members.

The residual deviation of the model is 1244.2 (with 1780 degrees of freedom), which is significantly lower than the deviation of the null model of 1854.6, indicating that the model fits well.
The final model realized the streamlining of variables while maintaining high explanatory power, and the AIC was reduced by about 4.6.

```{r}
#| label: check-dispersion
# Calculate the Pearson residual sum of squares
dispersion_ratio <- sum(residuals(selected_model1, type = "pearson")^2) / selected_model1$df.residual
cat("Dispersion ratio:", dispersion_ratio, "\n")

# Test for excessive discretisation
if (dispersion_ratio > 1.5) {
  cat("There is evidence of overdispersion. Consider using a Negative Binomial Model.\n")
} else {
  cat("No significant overdispersion detected. Poisson model is appropriate.\n")
}

```

In testing the final Poisson regression model for excessive dispersion, we used the ratio of the Pearson residual sum of squares to the residual degrees of freedom to calculate the dispersion ratio.
The results show dispersion ratio = 0.726, which is much lower than the commonly used determination threshold of 1.5, indicating that the model is not significantly over-dispersed.

Therefore, the Poisson model currently used is reasonable and there is no need to switch to more complex modeling methods such as negative binomial regression.
The model performs well in terms of both data fitting and assumptions, with stability and interpretability.

```{r}
#| label: model-evaluation
# Predicted fitted values
data$pred_poisson <- predict(selected_model1, type = "response")

# Plotting fitted values against true values
ggplot(data, aes(x = Total.Number.of.Family.members, y = pred_poisson)) +
  geom_point(alpha = 0.5, color = "blue") +
  ggtitle("Poisson Predictions vs. Actual Total.Number.of.Family.members") +
  xlab("Actual Total.Number.of.Family.members") +
  ylab("Predicted HTotal.Number.of.Family.members(Poisson)") +
  theme_minimal()

# Calculate the mean square error (MSE)
mse_poisson <- mean((data$Total.Number.of.Family.members - data$pred_poisson)^2)
cat("Mean Squared Error (MSE) of Poisson Model:", mse_poisson, "\n")
```

We evaluated the fitting effect of the filtered Poisson regression model.
Firstly, the predicted number of family members for each sample point was calculated by the predict() function, and the scatter plots of the actual and predicted values were plotted.
It can be seen from the plot that the predicted values show a good trend overall, especially in the interval of smaller number of family members (e.g., 2 to 8) the distribution is denser, and the predicted points are also roughly distributed along the diagonal direction, which indicates that the model has a strong fitting ability in most cases.

Further, the prediction accuracy was measured by the Mean Squared Error (MSE) metric, which was 3.36, indicating that the mean squared error between the model's predicted and actual values on each sample was 3.36 persons.
Although there are individual households with a high number of members (e.g., more than 12) with large deviations from the predicted values, the overall error level is within the acceptable range, and the model's prediction performance is stable.

```{r}
#| label: residuals-analysis
# Plotting residual distributions
ggplot(data.frame(resid_poisson = residuals(selected_model1, type = "pearson")),
       aes(x = resid_poisson)) +
  geom_histogram(bins = 30, fill = "pink", color = "black") +
  ggtitle("Residuals Distribution of Poisson Model") +
  xlab("Pearson Residuals") +
  theme_minimal()

# Standardised residuals vs. fitted values
plot(selected_model1$fitted.values, 
     residuals(selected_model1, type = "pearson"),
     xlab = "Fitted Values",
     ylab = "Pearson Residuals",
     main = "Residuals vs. Fitted Values")
abline(h = 0, col = "red", lwd = 2)
```

## Figure 1: Histogram of Pearson residual distribution

This figure shows the Pearson residual distribution of the Poisson regression model.
The overall residuals are approximately normally distributed, mainly concentrated around 0, indicating that most of the deviations between the predicted and actual values are small.
Although there is a slight long tail on the right side, no significant skewness or abnormal pattern is found.
This indicates that the model is fitted more accurately, there is no obvious systematic error, and the residual structure supports the basic assumptions of the model.

## Figure 2: Scatterplot of residuals vs. fitted values

This figure shows the relationship between the Pearson residuals and the fitted values of the model.
The residuals are roughly randomly distributed around the zero line, with no funnel-shaped distribution or curvilinear trend, indicating that the model does not suffer from severe heteroskedasticity or fitting bias.
The dense region of points is concentrated in the small and medium fitted value intervals, reflecting the characteristics of the data itself, but not destroying the model robustness.
Overall, the distribution of residuals is consistent with the assumptions of the Poisson regression model, further supporting the model's goodness of fit and suitability for explaining changes in the number of family members.

```{r}
# Negative binomial model fitting
nb_model <- glm.nb(
  Total.Number.of.Family.members ~
    Total.Household.Income +
    Total.Food.Expenditure +
    Household.Head.Age +
    Number.of.bedrooms +
    House.Floor.Area +
    House.Age +
    Type.of.Household +
    Electricity +
    Household.Head.Sex,
  data = data
)

selected_model2 <- step(poisson_full_model, direction = "backward", trace = TRUE)

summary(selected_model2)


```

In order to further verify the reasonableness of the model, we introduced the Negative Binomial Regression (NBR) model for comparative analysis after fitting the Poisson regression.
The reason is: the Poisson model assumes that the mean and variance of the response variable are equal, but there is often overdispersion in the actual data, i.e., the variance is larger than the mean, which will lead to underestimation of the standard error of the Poisson regression and exaggeration of the significance of the variable, thus affecting the accuracy of the inferred results.
Negative binomial regression can better handle such data characteristics by introducing additional discretization parameters.

In this study, we use glm.nb() to fit the full negative binomial model and optimize the variable selection by stepwise selection, and the final model is consistent with the structure of the Poisson simplified model.
The results show that the AIC value of the negative binomial model is 7115.5, which is exactly the same as the AIC of the Poisson model (also 7115.5), indicating that they are almost the same in terms of the trade-off between goodness-of-fit and complexity.

Combined with the previous excessive dispersion test (dispersion ratio = 0.726 \< 1.5) and residual plot analysis, the Poisson model did not show obvious dispersion problems, and the model residuals were well-distributed and stable, so we believe that the Poisson model is sufficient to explain the characteristics of the present data, and the structure of the model is more concise.

In addition, the estimated values of most key variables (e.g., household income, food expenditure, age of the household head, gender of the household head, etc.) in the two models are consistent with the significance levels, which further supports the robustness of the models.


